---
title       : Programming in R
subtitle    : A short introduction
author      : Alex Pacheco
job         : LTS Research Computing
logo        : lu.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
--- .rcr

## Read-And-Delete

1. Edit YAML front matter
2. Write using R Markdown
3. Use an empty line followed by three dashes to separate slides!

--- .rcr

## History

* R is a dialect of the S language
    - S was initiated at the Bell Labs as an internal statistical analysis environment.
    - Most well known implementation is S‐plus (most recent stable release was in 2010)
* R was first announced in 1993.
* The R core group was formed in 1997, who controls the source code of R (written in C)
* R 1.0.0 was released in 2000
* The current version is 3.1.3

--- .rcr

## Features

- R is a dialect of the S language
    - Language designed for statistical analysis
    - Similar syntax
- Available on most platform/OS
- Rich data analysis functionalities and sophisticated graphical capabilities
- Active development and very active community
    - CRAN: The Comprehensive R Archive Network
- Source code and binaries, user contributed packages and documentation
    - More than 6,000 packages available on CRAN as of last week
- Free to use

--- .rcr

## Running R

--- .rcr


## Getting Help

* Command Line

```
?<command name>
??<part of command name/topic>
```

--- .rcr

## Data Classes

* R has five atomic classes
* Numeric
    - Double is equivalent to numeric.
    - Numbers in R are treated as numeric unless specified otherwise.
* Integer
* Complex
* Character
* Logical
    - TRUE or FALSE
* You can convert data from one type to the other using the `as.<Type>` functions

--- .rcr


## Data Objects‐ Vectors

* Vectors can only contain elements of the same class
* Vectors can be constructed by
    -  Using the `c()` function (concatenate)
* Coercion will occur when mixed objects are passed to the `c()` function, as if the `as.<Type>()` function is explicitly called
    -  Using the `vector()` function
* One can use `[index]` to access individual element
    -  Indices start from 1

--- .rcr

## Examples

```{r}
# "#" indicates comment
# "<-" performs assignment operation (you can use "=" as well, but "<-" is preferred)
# numeric (double is the same as numeric)
d <- c(1,2,3)
# character
d <- c("1","2","3")
# you can covert at object with as.TYPE
# as. numeric changes the character vector created above to numeric
as.numeric(d)
# The conversion doesn't always work though
as.numeric("a")
```

--- .rcr

## Examples (contd)

```{r}
x <- c(0.5, 0.6) ## numeric
x <- c(TRUE, FALSE) ## logical
x <- c(T, F) ## logical
x <- c("a", "b", "c") ## character
# The ":" operator can be used to generate integer sequences
x <- 9:29 ## integer
x <- c(1+0i, 2+4i) ## complex
x <- vector("numeric", length = 10)
# Coercion will occur when objects of different classes are mixed
y <- c(1.7, "a") ## character
y <- c(TRUE, 2) ## numeric
y <- c("a", TRUE) ## character
# Can also coerce explicitly
x <- 0:6
class(x)
as.logical(x)
```

--- .rcr


## Vectorized Operations

* Lots of R operations process objects in a vectorized way
    -  more efficient, concise, and easier to read.

```{r}
x <- 1:4; y <- 6:9
x + y
x > 2
x * y
print( x[x >= 3] )
```

--- .rcr


## Data Objects - Matrices

* Matrices are vectors with a dimension attribute
* R matrices can be constructed
    -  Using the `matrix()` function
	  - Passing an dim attribute to a vector
	  -  Using the `cbind()` or `rbind()` functions
* R matrices are constructed column‐wise
* One can use `[<index>,<index>]` to access individual element

--- .rcr


## Example
```{r}
# Create a matrix using the matrix() function
m <- matrix(1:6, nrow = 2, ncol = 3)
m
dim(m)
attributes(m)
# Pass a dim attribute to a vector
m <- 1:10
m
dim(m) <- c(2, 5)
m
# Row binding and column binding
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
# Slicing
m <- 1:10
m
#m[c(1,2),c(2,4)]
```

--- .rcr


## Data Objects - Lists

* Lists are a special kind of vector that contains objects of different classes
* Lists can be constructed by using the `list()` function
* Lists can be indexed using `[[  ]]`

```{r}
# Use the list() function to construct a list
x <- list(1, "a", TRUE, 1 + 4i)
x
```

--- .rcr


## Data Objects - Data Frames
* Data frames are used to store tabular data
    -  They are a special type of list where every element of the list has to have the same length
    -  Each element of the list can be thought of as a column
    -  Data frames can store different classes of objects in each column
    -  Data frames also have a special attribute called `row.names`
    -  Data frames are usually created by calling `read.table()` or `read.csv()`
          -  More on this later
    -  Can be converted to a matrix by calling `data.matrix()`

--- .rcr


## Names

* R objects can have names

```{r}
# Each element in a vector can have a name
x <- 1:3
names(x)
names(x) <- c("a","b","c")
names(x)
x
```

--- .rcr

## Names (contd)

```{r}
# Lists
x <- list(a = 1, b = 2, c = 3)
x
# Names can be used to refer to individual element
x$a
```

--- .rcr

## Names (contd)

```{r}
# Columns and rows of matrices
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

--- .rcr


## Querying Object Attributes

* The `class()` function
* The `str()` function
* The `attributes()` function reveals attributes of an object (does not work with vectors)
    -  Class
    -  Names
    -  Dimensions
    -  Length
    -  User defined attributes
* They work on all objects (including functions)

--- .rcr

## Example

```{r}
m <- matrix(1:10, nrow = 2, ncol = 5)
str(matrix)
str(m)
str(matrix)
str(str)
```

--- .rcr


## Data Class - Factors

* Factors are used to represent categorical data.
* Factors can be unordered or ordered.
* Factors are treated specially by modelling functions like `lm()` and `glm()`

```{r}
# Use the factor() function to construct a vector of factors
# The order of levels can be set by the levels keyword
x <- factor(c("yes", "yes", "no", "yes", "no"), levels = c("yes", "no"))
x
```

--- .rcr


## Date and Time

* R has a Date class for date data while times are represented by POSIX formats
* One can convert a text string to date using the `as.Date()` function
* The `strptime()` function can deal with dates and times in different formats.
* The package "`lubridate`" provides many additional and convenient features

```{r}
# Dates are stored internally as the number of days since 1970-01-01
x <- as.Date("1970-01-01")
x
as.numeric(x)
x+1
```

--- .rcr

## Data and Time (contd)

```{r}
# Times are stored internally as the number of seconds since 1970-01-01
x <- Sys.time() ; x
as.numeric(x)
p <- as.POSIXlt(x)
names(unclass(p))
p$sec
```

--- .rcr


## Missing Values

* Missing values are denoted by `NA` or `NaN` for undefined mathematical operations.
    - `is.na()` is used to test objects if they are `NA`
    - `is.nan()` is used to test for `NaN`
    - `NA` values have a class also, so there are integer `NA`, character `NA`, etc.
    - A `NaN` value is also `NA` but the converse is not true

```{r}
x <- c(1,2, NA, 10,3)
is.na(x)
is.nan(x)
```

--- .rcr

## Missing Values (contd)

```{r}
x <- c(1,2, NaN, NA,4)
is.na(x)
is.nan(x)
```

--- .rcr

## Arithmetic Functions

<table>
<tr><th>Function</th><th>Description</th></tr>   
<tr><td><code>exp()</code></td><td>     Exponentiation</td></tr>
<tr><td><code>log()</code></td><td>     Natural Logarithm</td></tr>
<tr><td><code>log10()</code></td><td>   Logarithm to base 10</td></tr>
<tr><td><code>sqrt()</code></td><td>    square root</td></tr>
<tr><td><code>abs()</code></td><td>     absolute value</td></tr>
<tr><td><code>sin()</code></td><td>     sine</td></tr>
<tr><td><code>cos()</code></td><td>     cosine</td></tr>
<tr><td><code>floor()</code></td><td>    </td></tr>
<tr><td><code>ceiling()</code></td><td> rounding of numbers</td></tr>
<tr><td><code>round()</code></td><td>   </td></tr>
</table>

--- .rcr

## Simple Statistic Functions

<table border="1">
<tr><th>Function</th><th>Description</th></tr>
<tr><td><code>min()</code></td><td>minimum value</td></tr>
<tr><td><code>max()</code></td><td>maximum value</td></tr>
<tr><td><code>which.min()</code></td><td>location of minimum</td></tr>
<tr><td><code>which.max()</code></td><td>location of maximum</td></tr>
<tr><td><code>pmin()</code></td><td>elementwise minima of several vectors</td></tr>
<tr><td><code>pmax()</code></td><td>elementwise maxima of several vectors</td></tr>
<tr><td><code>sum()</code></td><td>sum of elements of a vector</td></tr>
<tr><td><code>mean()</code></td><td>mean of elements of a vector</td></tr>
<tr><td><code>prod()</code></td><td>products of elements of a vector</td></tr>
</table>

--- .rcr


## Distributions and Random Variables

* For each distribution R provides four functions: density (`d`), cumulative density (`p`), quantile (`q`), and random generation (`r`)
    - The function name is of the form `[d|p|q|r]<name of distribution>`
    - e.g. `qbinom()` gives the quantile of a binomial distribution


<table>
<tr><th>Distribution</th><th>Distribution name in R</th></tr>
<tr><td>Uniform</td><td><code>unif</code></td></tr>
<tr><td>Binomial</td><td><code>binom</code></td></tr>
<tr><td>Poisson</td><td><code>pois</code></td></tr>
<tr><td>Geometric</td><td><code>geom</code></td></tr>
<tr><td>Gamma</td><td><code>gamma</code></td></tr>
<tr><td>Normal</td><td><code>norm</code></td></tr>
<tr><td>Log Normal</td><td><code>lnorm</code></td></tr>
<tr><td>Exponential</td><td><code>exp</code></td></tr>
<tr><td>Student’s t</td><td><code>t</code></td></tr>
</table>

--- .rcr

## Examples: Distributions and Random Variables

```{r}
# Random generation from a uniform distribution.
runif(10, 2, 4)
# You can name the arguments in the function call.
runif(10, min = 2, max = 4)
# Given p value and degree of freedom, find the t-value.
qt(p=0.975, df = 8)
# The inverse of the above function call
pt(2.306, df = 8)
```

--- .rcr

## User Defined Functions
* Similar to other languages, functions in R are defined by using the `function()` directives
* The return value is the last expression in the function body to be evaluated.
* Functions can be nested
* Functions are R objects
    - For example, they can be passed as an argument to other functions

--- .rcr

## Control Structures
* Control structures allow one to control the flow of execution.

<table>
<tr><td><code>if … else</code></td><td>testing a condition</td></tr>
<tr><td><code>for</code></td><td>executing a loop (with fixed number of iterations)</td></tr>
<tr><td><code>while</code></td><td>executing a loop when a condition is true</td></tr>
<tr><td><code>repeat</code></td><td>executing an infinite loop</td></tr>
<tr><td><code>break</code></td><td>breaking the execution of a loop</td></tr>
<tr><td><code>next</code></td><td>skipping to next iteration</td></tr>
<tr><td><code>return</code></td><td>exit a function</td></tr>
</table>

--- .rcr


## Testing conditions

```{r eval=FALSE}
# Comparisons: <,<=,>,>=,==,!=
# Logical operations: !, &&, ||
if(x > 3 && x < 5) {
  print ("x is between 3 and 5")
} else if(x <= 3) {
  print ("x is less or equal to 3")
} else {
  print ("x is greater or equal to 5")
}
```

--- .rcr

## For Loops

```{r}
x <- c("a", "b", "c", "d")
# These loops have the same effect
# Loop through the indices
for(i in 1:4) {
  print(x[i])
}
# Loop using the seq_along() function
for(i in seq_along(x)) {
  print(x[i])
}
```

--- .rcr

## For Loops (contd)

```{r}
# Loop through the name
for(letter in x) {
  print(letter)
}
for(i in 1:4) print(x[i])
```

--- .rcr

## The apply Function
* The `apply()` function evaluate a function over
the margins of an array
    - More concise than the for loops (not necessarily
faster)

```{r}
# X: array objects
# MARGIN: a vector giving the subscripts which the function will be applied over
# FUN: a function to be applied
str(apply)
x <- matrix(rnorm(200), 20, 10)
# Row means
apply(x, 1, mean)
```

--- .rcr

## The apply Function (contd)

```{r}
# Column sums
apply(x, 2, sum)
# 25th and 75th Quantiles for rows
apply(x, 1, quantile, probs = c(0.25, 0.75))
```

--- .rcr

## The apply Function (contd)

```{r}
dim(x)
# Change the dimensions of x
dim(x) <- c(2,2,50)
# Take average over the first two dimensions
apply(x, c(1, 2), mean)
rowMeans(x, dims = 2)
```

--- .rcr

## Other Apply Functions
* `lapply`: Loop over a list and evaluate a function on each element
* `sapply`: Same as lapply but try to simplify the result
* `tapply`: Apply a function over subsets of a vector
* `mapply`: Multivariate version of lapply

--- .rcr

## Plyr Package
* In data analysis you often need to 
    - split up a big data structure into homogeneous pieces,
    - apply a function to each piece and then
    - combine all the results back together
* This split‐apply‐combine procedure is what
the `plyr` package is for.

--- .rcr

## Example: Plyr Package

```{r}
library(ggplot2)
library(plyr)
str(mpg)
str(ddply)
```

--- .rcr

## Example (contd)

```{r}
ddply(mpg, "cyl", summarise, mean = mean(cty))
```

--- .rcr

## Reading and Writing Data
* R understands many different data formats and has lots of ways of reading/writing them

<table>
<tr><th>Funtion (read)</th><th>Function (write)</th><th>Description</th></tr>
<tr><td><code>read.table</code></td><td><code>write.table</code></td><td></td></tr>
<tr><td><code>read.csv</code></td><td><code>write.csv</code></td><td>reading/writing tabular data</td></tr>
<tr><td><code>readLines</code></td><td><code>writeLines</code></td><td>reading/writing lines of a text file</td></tr>
<tr><td><code>source</code></td><td><code>dump</code></td><td>reading/writing in R code files</td></tr>
<tr><td><code>dget</code></td><td><code>dput</code></td><td>reading/writing in R code files</td></tr>
<tr><td><code>load</code></td><td><code>save</code></td><td>reading in/saving workspaces</td></tr>
<tr><td><code>unserialize</code></td><td><code>serialize</code></td><td>reading/writing single R objects in binary form</td></tr>
</table>

--- .rcr

## Reading Data with read.table

```{r}
str(read.table)
```

* `file` ‐ the name of a file, or a connection
* `header` ‐ logical indicating if the file has a header line
* `sep` ‐ a string indicating how the columns are separated
* `colClasses` ‐ a character vector indicating the class of each column in the dataset
* `nrows` ‐ the number of rows in the dataset
* `comment.char` ‐ a character string indicating the comment character
* `skip` ‐ the number of lines to skip from the beginning
* `stringsAsFactors` ‐ should character variables be coded as factors?

--- .rcr

## Reading Data with read.table (contd)

* The function will
    - Skip lines that begin with a #
    - Figure out how many rows there are (and how much memory needs to be allocated)
    - Figure out what type of variable is in each column of the table
* Telling R all these things directly makes R run faster and more efficiently.
* `read.csv()` is identical to `read.table()` except that the default separator is a comma.

--- .rcr

## Graphics in R
* There are three plotting systems in R
    - Base
* Convenient, but hard to adjust after the plot is created
    - Lattice
* Good for creating conditioning plot
    - Ggplot2
* Powerful and flexible, many tunable feature, may require some time to master
* Each has its pros and cons, so it is up to the users which one to choose

--- .rcr

## Graphics - Base
```{r}
summary(mpg)
```

--- .rcr

## Graphics - Base (contd)

```{r}
plot(hwy ~ displ, data=mpg)
```

--- .rcr

## Graphics - Base (contd)

```{r}
boxplot(hwy ~ class, data = mpg, xlab = "Class", ylab = "High Way Mileage")
```

--- .rcr

## Graphics - Lattice

```{r}
mpg <- transform(mpg, year = factor(year)) 
xyplot(hwy ~ displ | year*drv, mpg, layout = c(6,1))
```

--- .rcr

## Graphics - ggplot2
```{r}
qplot(displ, cty, data = mpg, color = drv)
```

--- .rcr

## Graphics - ggplot2
```{r}
qplot(displ, hwy, data = mpg, facets = drv ~ year)
```

```{r eval=FALSE}
ggplot(emiByType, aes(year, sum, colour = type)) +
 geom_point(size = 5) +
 geom_line(size = 1) +
 ggtitle("Total Emission in Baltimore City by Source Type") +
 labs(x = "Year", y = "Total Emissions (ton)", fontsize = 20) +
 theme_economist() + scale_colour_economist() +
 theme(axis.title=element_text(size=14,face="bold"), legend.title =
   element_text(size = rel(1.5), face = "bold"))
```

--- .rcr &twocol

## Histograms

*** =left

```{r eval=F}
library(RColorBrewer)
data(VADeaths)
par(mfrow=c(2,3))
hist(VADeaths,breaks=10, col=brewer.pal(3,"Set3"),main="Set3 3 colors")
hist(VADeaths,breaks=3 ,col=brewer.pal(3,"Set2"),main="Set2 3 colors")
hist(VADeaths,breaks=7, col=brewer.pal(3,"Set1"),main="Set1 3 colors")
hist(VADeaths,,breaks= 2, col=brewer.pal(8,"Set3"),main="Set3 8 colors")
hist(VADeaths,col=brewer.pal(8,"Greys"),main="Greys 8 colors")
hist(VADeaths,col=brewer.pal(8,"Greens"),main="Greens 8 colors")
```

*** =right

```{r echo=F}
library(RColorBrewer)
data(VADeaths)
par(mfrow=c(2,3))
hist(VADeaths,breaks=10, col=brewer.pal(3,"Set3"),main="Set3 3 colors")
hist(VADeaths,breaks=3 ,col=brewer.pal(3,"Set2"),main="Set2 3 colors")
hist(VADeaths,breaks=7, col=brewer.pal(3,"Set1"),main="Set1 3 colors")
hist(VADeaths,,breaks= 2, col=brewer.pal(8,"Set3"),main="Set3 8 colors")
hist(VADeaths,col=brewer.pal(8,"Greys"),main="Greys 8 colors")
hist(VADeaths,col=brewer.pal(8,"Greens"),main="Greens 8 colors")
```

***

--- .rcr

## Line Chart

```{r}
plot(AirPassengers,type="l")  #Simple Line Plot
```

--- .rcr &twocol_width

## Bar Chart

*** =left width:48%
```{r}
barplot(iris$Petal.Length) #Creating simple Bar Graph
```
*** =right width:48%
```{r}
barplot(iris$Sepal.Length,col  = brewer.pal(3,"Set1"))
```
***

--- .rcr
## Bar Chart
```{r}
barplot(table(iris$Species,iris$Sepal.Length),col  = brewer.pal(3,"Set1")) #Stacked Plot
```

--- .rcr

## Box Plot
```{r}
boxplot(iris$Petal.Length~iris$Species) #Creating Box Plot between two variable
```

--- .rcr &twocol_width

## Box Plot

*** =left width:48%

```{r eval=F}
data(iris)
par(mfrow=c(2,2))
boxplot(iris$Sepal.Length,col="red")
boxplot(iris$Sepal.Length~iris$Species,col="red")
boxplot(iris$Sepal.Length~iris$Species,col=heat.colors(3))
boxplot(iris$Sepal.Length~iris$Species,col=topo.colors(3))
```

*** =right width:48%

```{r echo=F}
data(iris)
par(mfrow=c(2,2))
boxplot(iris$Sepal.Length,col="red")
boxplot(iris$Sepal.Length~iris$Species,col="red")
boxplot(iris$Sepal.Length~iris$Species,col=heat.colors(3))
boxplot(iris$Sepal.Length~iris$Species,col=topo.colors(3))
```

***

--- .rcr

## Scatter Plots
```{r}
plot(x=iris$Petal.Length) #Simple Scatter Plot
plot(x=iris$Petal.Length,y=iris$Species) #Multivariate Scatter Plot
plot(iris,col=brewer.pal(3,"Set1"))
pie(table(iris$Species))
```

--- .rcr &twocol_width

## Hexbin Binning

* We can use the hexbin package in case we have multiple points in the same place (overplotting). Hexagon binning is a form of bivariate histogram useful for visualizing the structure in datasets with large n. 

*** =left width:48%

```{r eval=F}
#library(ggplot2) to get diamonds dataset
library(hexbin)
a=hexbin(diamonds$price,diamonds$carat,xbins=40)
library(RColorBrewer)
plot(a)
```

*** =right width:48%

```{r, fig.height = 6, fig.align = 'center', echo=F}
#library(ggplot2) to get diamonds dataset
library(hexbin)
a=hexbin(diamonds$price,diamonds$carat,xbins=40)
library(RColorBrewer)
plot(a)
```

***

--- .rcr 

## (Contd)

```{r, fig.height = 6, fig.align = 'center', warning=FALSE}
# We can also create a color palette and then use the hexbin plot function for a better visual effect. 
rf <- colorRampPalette(rev(brewer.pal(40,'Set3')))
hexbinplot(diamonds$price~diamonds$carat, data=diamonds, colramp=rf)
```

--- .rcr

## Mosaic Plot

* A mosaic plot can be used for plotting categorical data very effectively with the area of the data showing the relative proportions.
```{r, fig.height = 5.5, fig.align = 'center'}
data(HairEyeColor)
mosaicplot(HairEyeColor)
```

--- .rcr

## Heat Map

* Heat maps enable you to do exploratory data analysis with two dimensions as the axis and the third dimension shown by intensity of color. 
* However you need to convert the dataset to a matrix format. Here’s the code:

```{r , fig.height = 5, fig.align = 'center'}
heatmap(as.matrix(mtcars))
```

--- .rcr

## Heat Map (contd)

* You can use `image()` command also for this type of visualization as:
```{r, fig.height = 5, fig.align = 'center'}
image(as.matrix(mtcars[2:7]))
```

--- .rcr &twocol_width

## 3D Graphs

*** =left width:48%

* Install package R Commander 
`install.packages("Rcmdr")`
* Install Lattice package
* Correlogram (GUIs)

Correlogram help us visualize the data in correlation matrices. Here’s the code:

```{r}
cor(iris[1:4])
```

*** =right width:48%

```{r, fig.height = 5, fig.align = 'center'}
corrgram(iris)
```

***

--- .rcr


## Slide with R Code and Output

```{r}
summary(cars)
```

--- .rcr

## Slide with Plot

```{r, message=FALSE}
plot(cars)
```



